### 2013-02-15 doc/icml2013/algo.tex

%%% second moments
Intuitively, performing regression on only $y$ given $x$ provides only first-order
information.  The key insight is that we can perform regression
on higher order powers to obtain more information about the parameters.
Specifically, for an integer $p \ge 1$, let us define the average
$m$-th order tensor power of the parameters as follows:
\begin{align}
M_p &\eqdef \sum_{h=1}^k \pi_h \beta_h\tp{p}. \label{eqn:Mp} % \in (\Re^{d})\tp{p}.
\end{align}
Now consider performing regression on $y^2$ given $x\tp{2}$.
Expanding $y^2 = (\innerp{\beta_h}{x} + \epsilon)^2$,
using the fact that $\innerp{\beta_h}{x}^p = \innerp{\beta_h\tp{p}}{x\tp{p}}$,
we have:
\begin{align}
y^2 &= \innerp{M_2}{x\tp{2}} + \sigma^2 + \eta_2(x), \label{eqn:y2} \\
\eta_2(x) &= \innerp{\beta_h\tp{2} - M_2}{x\tp{2}} + 2 \epsilon \innerp{\beta_h\tp{2}}{x\tp{2}} + (\epsilon^2 - \sigma^2). \nonumber
\end{align}
Again, we have expressed $y^2$ has a linear function of $x\tp{2}$
with regression coefficients $M_2$, plus a known bias $\sigma^2$ and noise.
Importantly, the noise has mean zero; 
in fact each of the three terms has mean zero
by (i) definition of $M_2$, (ii) independence of $\epsilon$ and $h$,
and (iii) the fact that $\E[\epsilon^2] = \sigma^2$.

@@

% Unmixing
At a conceptual level, Spectral Experts bears some resemblence to the unmixing
algorithm for estimation of restricted PCFGs
\cite{hsu12identifiability}.
In that work, the observations (moments) provided a linear combination over
the compound parameters.  ``Unmixing'' involves solving for the compound
parameters by inverting a mixing matrix.  In this work,
each data point (appropriately transformed) provides a noisy slice of the
compound parameters; retrieving the compound parameters is more involved.

@@

\begin{algorithm}[t]
  \caption{Spectral Experts}
  \label{algo:spectral-experts}
  \begin{algorithmic}[1]
    \INPUT Data $\mathcal{D}_r = \{ (\xn{1}, \yn{1}), \cdots, (\xn{n}, \yn{n}) \}$, $r = 1, 2, 3$.
    \OUTPUT Parameters $\hat\theta = (\hat \pi, [\hat \beta_1 \mid \cdots \mid \hat \beta_k])$.
    \STATE Estimate compound parameters $M_2, M_3$ using \textbf{low-rank regression}:
    \begin{align}
      &\hat M_1 = \arg\min_{M_1} \label{eqn:estimateM1} \\
      &\quad\frac{1}{2n} \sum_{(x,y) \in \sD_1} (\innerp{M_1}{x\tp{2}} - y)^2, \nonumber \\
      &\hat M_2 = \arg\min_{M_2} \label{eqn:estimateM2} \\
      &\quad\frac{1}{2n} \sum_{(x,y) \in \sD_2} (\innerp{M_2}{x\tp{2}} + \sigma^2 - y^2)^2 + \lambda_n^{(2)} \|M_2\|_*, \nonumber \\
      &\hat M_3 = \arg\min_{M_3} \label{eqn:estimateM3} \\
      &\quad\frac{1}{2n} \sum_{(x,y) \in \sD_3} (\innerp{M_3}{x\tp{3}} + 3 \sigma^2\innerp{\hat M_1}{x} - y^3)^2 + \lambda_n^{(3)} \|M_3\|_*. \nonumber
    \end{align}
    \STATE Estimate the parameters $\theta = (\pi, B)$ using \textbf{tensor factorzation}:
    \begin{enumerate}
      \item [(a)] Compute whitening matrix $W \in \Re^{d \times k}$ (such that $W^\top
      \hat M_2 W = I$) using SVD.
      \item [(b)] Compute the eigenvalues $\{a_h\}_{h=1}^k$
      and eigenvectors $\{v_h\}_{h=1}^k$
      of the whitened tensor $M_3(W, W, W) \in \Re^{k \times k \times k}$
      by using the robust tensor power method.
    \item [(c)] Set parameter estimates $\hat\pi_h = \frac{1}{a_h^2}$
    and $\hat\beta_h = (W^{\top})^\dagger (a_h v_h)$.
    \end{enumerate}
  \end{algorithmic}
\end{algorithm}

@@

In this section, we conduct a theoretical analysis of Spectral Experts.
In particular, we prove that it enjoys consistent parameter estimates.
\begin{theorem}[Convergence of Spectral Experts]
Each $\sD_p$ (for $p = 1, 2, 3$) consists of $n$ i.i.d. points drawn from the mixture
of linear regressions model with parameter $\theta^*$
(having three independent copies simplifies the analysis).
Assume that $\E[\unfold(x\tp{2})\tp{2}] \succ 0$ 
and $\E[\unfold(x\tp{3})\tp{2}] \succ 0$.
Let $\|x\|_2 \le R$ with probability $1$
and let $\|\beta_h\|_2 \le L$ for all $h \in [k]$.
There exists universal constants $c_1,c_2$ such that
if the number of samples is
$n \ge \frac{c_1 k d^2 L R}{\epsilon}$
and $\lambda_n^{(2)} \ge \frac{c_2 L R}{\sqrt{n}}$,
and $\lambda_n^{(3)} \ge \frac{c_3 L R}{\sqrt{n}}$,
then the parameter estimates $\hat\theta$ returned by
\algorithmref{algo:spectral-experts} (with the columns appropriate permuted)
satisfies 
$\|\hat\pi - \pi^*\|_{\infty} \le \frac{\epsilon}{???}$
and for all $h \in [k]$,
$\|\hat\beta_h - \beta^*_h\|_2 \le \epsilon ???$,
\end{theorem}
\begin{proof}
Let $T^* = M_3^*(W^*, W^*, W^*)$.
By Theorem 5.1 of \cite{AnandkumarGeHsu2012},
Need to convert error from $\epsilon$
hit by $(W^\top)^\dagger$
\end{proof}

@@

\subsection{Identifiability}

In ordinary linear regression, the regression coefficients are identifiable
if and only if the data has full rank: $\E[x\tp{2}] \succ 0$.
However, in mixture of linear regressions, this is insufficient.  
For example, suppose $x = (1, t, \cdots, t^p)$, 
the common polynomial basis expansion.
Regressing on all powers up to $p$ yield at most $3$.

>>>>

%%% second moments
Intuitively, performing regression on only $y$ given $x$ provides only first-order
information.  The key insight is that we can perform regression
on higher order powers to obtain more information about the parameters.
Specifically, for an integer $p \ge 1$, let us define the average
$m$-th order tensor power of the parameters as follows:
\begin{align}
M_p &\eqdef \sum_{h=1}^k \pi_h \beta_h\tp{p}. \label{eqn:Mp} % \in (\Re^{d})\tp{p}.
\end{align}
Now consider performing regression on $y^2$ given $x\tp{2}$.
Expanding $y^2 = (\innerp{\beta_h}{x} + \epsilon)^2$,
using the fact that $\innerp{\beta_h}{x}^p = \innerp{\beta_h\tp{p}}{x\tp{p}}$,
we have:
\begin{align}
y^2 &= \innerp{M_2}{x\tp{2}} + \sigma^2 + \eta_2(x), \label{eqn:y2} \\
\eta_2(x) &= \innerp{\beta_h\tp{2} - M_2}{x\tp{2}} + 2 \epsilon \innerp{\beta_h\tp{2}}{x\tp{2}} + (\epsilon^2 - \sigma^2). \nonumber
\end{align}
Again, we have expressed $y^2$ has a linear function of $x\tp{2}$
with regression coefficients $M_2$, plus a known bias $\sigma^2$ and noise.\footnote{If $\sigma^2$ were not known,
we could treat it as another coefficient
to be estimated.  The coefficients $M_2$ and $\sigma^2$ can be estimated jointly
provided that $x$ does not already contain a bias ($x_j$ must be non-constant for every $j \in [d]$).}
Importantly, the noise has mean zero; 
in fact each of the three terms has mean zero
by (i) definition of $M_2$, (ii) independence of $\epsilon$ and $h$,
and (iii) the fact that $\E[\epsilon^2] = \sigma^2$.

@@

\paragraph{Remarks}

% Spectral
In recent years, there has a been substantial interest in ``spectral'' methods
for learning latent-variable models.  One line of work has
focused on observerable operator models \cite{hsu09spectral}
in which the true parameters are not recovered, but a reparametrization is used
which enables prediction.
Another line of work is based on the method of moments and uses eigendecomposition of a certain tensor
to recover the parameters \cite{anandkumar12moments,anandkumar12svd,AnandkumarHsuKakade2012}.
Our work extends this line of work into the space of discriminative models,
which requires regression to obtain the desired tensor.

% Unmixing
In spirit, Spectral Experts also bears some resemblence to the unmixing
algorithm for estimation of restricted PCFGs
\cite{hsu12identifiability}.
In that work, the observations (moments) provided a linear combination over
the compound parameters.  ``Unmixing'' involves solving for the compound
parameters by inverting a mixing matrix.  In this work,
each data point (appropriately transformed) provides a noisy slice of the
compound parameters; retrieving the compound parameters is more involved.

@@

\begin{algorithm}[t]
  \caption{Spectral Experts}
  \label{algo:spectral-experts}
  \begin{algorithmic}[1]
    \INPUT Three independent datasets $\mathcal{D}_r = \{ (\xn{1}, \yn{1}), \cdots, (\xn{n}, \yn{n}) \}$ for $r = 1, 2, 3$;
    regularization strengths $\lambda_n^{(2)} = \frac{c_2}{\sqrt{n}}$, $\lambda_n^{(3)} = \frac{c_3}{\sqrt{n}}$;
    observation noise variance $\sigma^2$.
    \OUTPUT Parameters $\hat\theta = (\hat \pi, [\hat \beta_1 \mid \cdots \mid \hat \beta_k])$.
    \STATE Estimate compound parameters $M_2, M_3$ using \textbf{low-rank regression}:
    \begin{align}
      &\hat M_1 = \arg\min_{M_1} \label{eqn:estimateM1} \\
      &\quad\quad\frac{1}{2n} \sum_{(x,y) \in \sD_1} (\innerp{M_1}{x} - y)^2, \nonumber \\
      &\hat M_2 = \arg\min_{M_2} \quad \lambda_n^{(2)} \|M_2\|_* + \label{eqn:estimateM2} \\
      &\quad\quad\frac{1}{2n} \sum_{(x,y) \in \sD_2} (\innerp{M_2}{x\tp{2}} + \sigma^2 - y^2)^2, \nonumber \\
      &\hat M_3 = \arg\min_{M_3} \quad \lambda_n^{(3)} \|M_3\|_* + \label{eqn:estimateM3} \\
      &\quad\quad\frac{1}{2n} \sum_{(x,y) \in \sD_3} (\innerp{M_3}{x\tp{3}} + 3 \sigma^2\innerp{\hat M_1}{x} - y^3)^2. \nonumber
    \end{align}
    \STATE Estimate the parameters $\theta = (\pi, B)$ using \textbf{tensor factorzation}:
    \begin{enumerate}
      \item [(a)] Compute whitening matrix $\hat W \in \Re^{d \times k}$ (such that $\hat W^\top
      \hat M_2 \hat W = I$) using SVD.
      \item [(b)] Compute the eigenvalues $\{\hat a_h\}_{h=1}^k$
      and eigenvectors $\{\hat v_h\}_{h=1}^k$
      of the whitened tensor $\hat M_3(\hat W, \hat W, \hat W) \in \Re^{k \times k \times k}$
      by using the robust tensor power method.
    \item [(c)] Set parameter estimates $\hat\pi_h = \frac{1}{\hat a_h^2}$
    and $\hat\beta_h = (\hat W^{\top})^\dagger (\hat a_h \hat v_h)$.
    \end{enumerate}
  \end{algorithmic}
\end{algorithm}

@@

%Define $\sM_p = \{ \sum_{h=1}^k \pi_h \beta_h\tp{p} : \pi_h \in \R^k, \beta_h
%\in \R^d \}$ be the set of all $p$-th order compound parameters.
%Let $\bM_p$ be the set of

%Define the symmetric rank
%\begin{align}
%\E[\cvec(x\tp{p})\tp{2}] \succ 0.
%\end{align}

In this section, we conduct a theoretical analysis of Spectral Experts.
In particular, we prove that it enjoys consistent parameter estimates.
\begin{theorem}[Convergence of Spectral Experts]
  \label{thm:convergence}
Each $\sD_p$ (for $p = 1, 2, 3$) consists of $n$ i.i.d. points drawn from the mixture
of linear regressions model with parameter $\theta^*$
(having three independent copies simplifies the analysis).
Assume that $\E[\cvec(x\tp{p})\tp{2}] \succ 0$ for each $p \in \{1,2,3\}$.
Assume $d \ge k$.
Let $\|x\|_2 \le R$ with probability $1$
and let $\|\beta_h\|_2 \le L$ for all $h \in [k]$.
There exists universal constants $c_1,c_2$ such that
if the number of samples is
$n \ge \frac{c_1 k d^2 L R}{\epsilon}$
and $\lambda_n^{(2)} \ge \frac{c_2 L R}{\sqrt{n}}$,
and $\lambda_n^{(3)} \ge \frac{c_3 L R}{\sqrt{n}}$,
then the parameter estimates $\hat\theta$ returned by
\algorithmref{algo:spectral-experts} (with the columns appropriate permuted)
satisfies 
$\|\hat\pi - \pi^*\|_{\infty} \le \frac{\epsilon}{???}$
and for all $h \in [k]$,
$\|\hat\beta_h - \beta^*_h\|_2 \le \epsilon ???$,
\end{theorem}
\begin{proof}
Let $T^* = M_3^*(W^*, W^*, W^*)$.
By Theorem 5.1 of \cite{AnandkumarGeHsu2012},
Need to convert error from $\epsilon$
hit by $(W^\top)^\dagger$
\end{proof}

@@

\subsection{Identifiability from moments}

In ordinary linear regression, the regression coefficients $\beta \in \Re^d$ are
identifiable if and only if the data has full
rank: $\E[x\tp{2}] \succ 0$,
and furthermore, identifying $\beta$ requires only moments
$\E[xy]$ and $\E[x\tp{2}]$ (by observing the optimality conditions for \refeqn{y1}).
However, in mixture of linear regressions, these two moments only allow us to recover $M_1$.
\refthm{convergence} shows that if we have the higher order analogues,
$\E[x\tp{p}y\tp{p}]$ and $\E[x\tp{2p}]$ for $p \in \{1,2,3\}$,
we can then identify the parameters $\theta = (\pi, B)$,
provided $\E[\cvec(x\tp{p})\tp{2}] \succ 0$ for $p \in \{1,2,3\}$.

We note that the positive definite constraints are more stringent.
For example, suppose $x = (1, t, t^2)$,
the common polynomial basis expansion, so that all the coordinates are deterministically related.
Then $\E[x\tp{2}] \succ 0$ can easily be satisfied, say with $x \sim \normal{0}{I}$.
However, one can verify that $\E[\cvec(x\tp{2})\tp{2}]$ is singular,
since $\cvec(x\tp{2}) = [1 \cdot 1, t\cdot t, 2(1 \cdot t^2), 2(t \cdot t^2), (t^2 \cdot t^2)]$ contains
components $t \cdot t$ and $2(1 \cdot t^2)$, which are linearly dependent.
So our method would not be able to identify the parameters of a mixture of
linear regressions for this distribution over $x$.

We now show that some amount of unidentifiability is intrinsic to the model,
not just our estimation procedure.
Suppose $x = (t, \dots, t^d)$.
Even if we observed all moments
$\E[x\tp{p}y\tp{p}]$ and $\E[x\tp{2p}]$ for $p \in [r]$,
all the coordinates would be monomials up to degree $2dr$,
the therefore the moments live in a $2(dr)$-dimensional subspace.
The parameters live in a subspace of at least dimension $dk$.
Therefore, at least $r \ge k/2$ moments are required for identifiability of any
algorithm for this monomial example.

%In fact, considering moments up to power $p
%Regressing on all powers up to $p$ yield at most $3$.

### 2013-02-15 doc/icml2013/intro.tex

We use $x\tp{p}$ to represent the $p$-th order tensor
formed by taking the outer product of $x \in \Re^d$; i.e. $x\tp{p}_{i_1 i_2 \ldots
i_p} = x_{i_1} x_{i_2} \cdots x_{i_p}$. We will use
$\innerp{\cdot}{\cdot}$ to denote the generalized dot product between two $p$-th
order tensors: $\innerp{X}{Y} = \sum_{i_1, i_2, \ldots i_p} X_{i_1,
i_2, \ldots i_p} Y_{i_1, i_2, \ldots i_p}.$
For a $p$-th order tensor $X \in (\Re^d)\tp{p}$, let $X_{(i)} \in \Re^{d \times d^{p-1}}$
be the mode-$i$ unfolding of $X$.
%For $X \in (\Re^d)\tp{p}$ and $Y \in (\Re^d)\tp{(p+q)}$,% $p, q > 0$,
%$\innerp{X}{Y} \in \Re^{d^q}$ is the projection of $X$ onto the first
%$p$ modes of $Y$, i.e. $\innerp{X}{Y}_{i_1, \ldots, i_q}
%= \sum_{j_1, \ldots, j_p} X_{j_1, \ldots, j_p} Y_{j_1, \ldots, j_p, i_1,
%\ldots, i_q}$.

% Norms
For a matrix $X$,
let $\|X\|_*$ denote the nuclear (trace) norm (sum of singular values),
let $\|X\|_F$ denote the Frobenius norm (square root of sum of squares of singular values),
let $\|X\|_\text{op}$ denote the operator norm (largest singular value).
For a rank-$K$ tensor $X$,
let $\|X\|_* = \frac{1}{p} \sum_{i=1}^p \|X_{(i)}\|_*$ denote
the average nuclear norm over all $p$ unfoldings,
and let $\|X\|_\text{op} = \frac{1}{p} \sum_{i=1}^p \|X_{(i)}\|_\text{op}$
denote the average operator norm over all $p$ unfoldings.
Let $\unfold(X)$ be the vectorized version of $X$.

@@

>>>>

@@

We use $x\tp{p}$ to represent the $p$-th order tensor
formed by taking the outer product of $x \in \Re^d$; i.e. $x\tp{p}_{i_1 \ldots
i_p} = x_{i_1} \cdots x_{i_p}$. We will use
$\innerp{\cdot}{\cdot}$ to denote the generalized dot product between two $p$-th
order tensors: $\innerp{X}{Y} = \sum_{i_1, \ldots i_p} X_{i_1,
\ldots i_p} Y_{i_1, \ldots i_p}.$
A tensor $X$ is symmetric if for all $i,j \in [d]^p$ which are permutations,
$X_{i_1 \cdots i_p}$ = $X_{j_1 \cdots j_p}$ (all tensors in this paper will be symmetric).
For a $p$-th order tensor $X \in (\Re^d)\tp{p}$, let $X_{(i)} \in \Re^{d \times d^{p-1}}$
be the mode-$i$ unfolding of $X$.
%For $X \in (\Re^d)\tp{p}$ and $Y \in (\Re^d)\tp{(p+q)}$,% $p, q > 0$,
%$\innerp{X}{Y} \in \Re^{d^q}$ is the projection of $X$ onto the first
%$p$ modes of $Y$, i.e. $\innerp{X}{Y}_{i_1, \ldots, i_q}
%= \sum_{j_1, \ldots, j_p} X_{j_1, \ldots, j_p} Y_{j_1, \ldots, j_p, i_1,
%\ldots, i_q}$.

% Norms
For a matrix $X$,
let $\|X\|_*$ denote the nuclear (trace) norm (sum of singular values),
let $\|X\|_F$ denote the Frobenius norm (square root of sum of squares of singular values),
let $\|X\|_\text{op}$ denote the operator norm (largest singular value).
For a rank-$K$ tensor $X$,
let $\|X\|_* = \frac{1}{p} \sum_{i=1}^p \|X_{(i)}\|_*$ denote
the average nuclear norm over all $p$ unfoldings,
and let $\|X\|_\text{op} = \frac{1}{p} \sum_{i=1}^p \|X_{(i)}\|_\text{op}$
denote the average operator norm over all $p$ unfoldings.

For a tensor $X \in (\Re^d)\tp{p}$,
let $\cvec(X) \in \Re^{\sum_{i=1}^p \binom{n}{i}}$ be
the collapsed vectorization of $X$: each component of $\cvec(X)$
is indexed by a vector of counts $(c_1, \dots, c_d)$ with total sum $\sum_i c_i = p$.
The value of that component is
$\sum_{k \in K(c)} X_{k_1 \cdots k_p}$,
where $K(c) = \{ k \in [d]^p : \forall i \in [d], c_i = |\{ j \in [p] : k_j = i \}| \}$ are the set of index vectors $k$ with that count profile.
For example, for $X \in \Re^{d \times d}$,
$\cvec(X) = (X_{ii} : i \in [d]; X_{ij} + X_{ji} : i,j \in [d], i<j)$.
For symmetric tensors, $\innerp{X}{Y} = \innerp{\cvec(X)}{\cvec(Y)}$.
